\chapter{Mathematical Formulation}
\label{chap:mipformulation}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This chapter describes a method for solving the problem which is purely based on mixed integer programming (MIP).

This approach divides the problem in two modules, which were called 'tactical' and 'operational' module.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Integer programming}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mixed integer program}

Introduces MIP concept and general formulation.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Linear relaxation}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Integrality gap}

Introduces integrality gap concept.

As in all combinatorial scheduling models, the problem grows more complex as the number of side constraints increases.\fixme{Colocar essa frase no lugar adequado!}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Branch and Bound}

Introduces branch and bound concept.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{MIP Solver}

Introduces CPLEX and Gurobi.



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%    TATIC MODULE     %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Tactical module}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Notation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Sets}

\begin{itemize}
\item $Al$ - Set of students. Elements of $Al$ are called $a$.
\item $C$ - Set of majors. Elements of $C$ are called $c$.
\item $CP$ - Set of campi. Elements of $CP$ are called $cp$.
\item $D$ - Set of courses. Elements of $D$ are called $d$.
\item $D_{a}$ - Set of courses required by student $a$.
\item $H_{t}$ - Set of time slots of day $t$. Elements of $H_{t}$ are called $h$.
\item $I_{d}$ - Set of sections of a course $d$. Elements of $I_{d}$ are called $i$.
\item $O$ - Set of majors offers. Elements of $O$ are called $oft$.
\item $S_{u}$ - Set of classrooms of block $u$. Elements of $S_{u}$ are called $s$.
\item $T$ - Set of weekdays. Elements of $T$ are called $t$.
\item $U$ - Set of blocks. Elements of $U$ are called $u$.
\item $SL$ - Set of calenders. Elements of $SL$ are called $sl$.
\item $P$ - Set of all professors. Elements of $P$ are called $p$.
\item $PV$ - Set of virtual professors. Elements of $PV$ are called $pv$ or $p$.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Model Data}

\begin{itemize}
\item $Cap_{s}$ - Capacity of classroom $s$.
\item $nCreds_{d}$ - Total of credits of course $d$.
\item $M$ - big $M$.
\item $N_{d,k,t}$ - number of credits for course $d$ at day $t$ of credits split rule $k$.
\item $NCH_{d,hi,hf}$ - number of credits for course $d$ from time slot $hi$ to time slot $hf$.
\item $MinSize$ - minimum number of students required for offering a class.
\item $delta_{a,t}$ - maximum of idle time allowed at day $t$ for student $a$. Used only for gap-constraints, usually is the sum of intervals at day $t$ of the student's calender.
\item $delta_{t,f}$ - maximum of idle time allowed at phase $f$ of day $t$ for a professor. Used only for gap-constraints, usually is the average sum of intervals at phase $f$ of day $t$ of calenders.
\item $pv$ - Unique virtual professor, to be used whenever the set of real professors is not enough for demand satisfaction.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Variables}

\begin{itemize}
\item $v_{a,i,d,s,t,hi,hf}$ - binary variable, indicates if student $a$ is assigned to a class of section $i$ of course $d$, at classroom $s$, from time slot $hi$ to time slot $hf$ of day $t$. 
\item $x_{i,d,u,s,hi,hf,t}$ - binary variable, indicates if section $i$ of course $d$ is assigned to classroom $s$ from time slot $hi$ to time slot $hf$ of day $t$. 
\item $z_{i,d}$ - binary variable, indicates if section $i$ of course $d$ is offered. 
\item $o_{i,d,s}$ - binary variable, indicates if section $i$ of course $d$ is offered at classroom $s$. 
\item $fd_{d,a}$ - binary slack variable, indicates if student $a$ has his request for course $d$ not satisfied.
\item $fkm_{d,i,t}$ - slack variable for inferior limit for credits split rule constraint. 
\item $fkp_{d,i,t}$ - slack variable for superior limit for credits split rule constraint.
\item $m_{d,i,k}$ - binary variable, indicates if a credits split rule $k$ was chosen for section $i$ of course $d$.
\item $s_{i,d,a}$ - binary variable, indicates if student $a$ is assigned to section $i$ of course $d$.
\item $k_{p,i,d,u,h,t}$ - binary variable, indicates if professor $p$ teaches to section $i$ of course $d$ at block $u$ at time slot $h$ and day $t$.
\item $y_{p,i,d,cp}$ - binary variable, indicates if professor $p$ is assigned to section $i$ of course $d$ of campus $cp$.
\item $hip_{p,t,f}$ - integer variable, indicates the time in minutes of the first time slot assigned to professor $p$ at session $f$ of day $t$. For example, if the first class assigned to $p$ on $t=Monday$ and $f=morning$ starts at 9:30 am, then $hip_{p,t,f}=9\cdot 60 + 30 = 570$.
\item $hfp_{p,t,f}$ - integer variable, indicates the time in minutes of the end of the last time slot assigned to professor $p$ at session $f$ of day $t$.
\item $hia_{a,t}$ - integer variable, indicates the time in minutes of the first time slot assigned to student $a$ at day $t$.
\item $hfa_{a,t}$ - integer variable, indicates the time in minutes of the end of the last time slot assigned to student $a$ at day $t$.
\item $fpgap_{p,t,f}$ - integer slack variable, indicates the gap in session $f$ of day $t$ of professor $p$'s schedule.
\end{itemize}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{MIP  Assignment Formulation}


\subsubsection{Objective function}
$$
\begin{array}{rl}
   \mbox{MIN} &
			\sum\limits_{a \in A}\sum\limits_{d \in D} \cdot fd_{d,a}
      \\
      &
       + \sum\limits_{d \in D} 
\sum\limits_{t \in T} \sum\limits_{i \in I_{d}} (fkp_{d,i,k} + fkm_{d,i,k})
      \\
      &
      + \sum\limits_{p \in P} \sum\limits_{t \in T} \sum\limits_{f \in F} fpgap_{p,t,f}
      \\
      &
      + \sum\limits_{i \in I_{d}} \sum\limits_{d \in D} \sum\limits_{cp \in CP} y_{pv,i,d,cp}
\end{array}
$$


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\subsubsection{Assigns 'v' to 'x' and ensures classroom capacity}
\begin{eqnarray}
\sum\limits_{a \in A} v_{a,i,d,s,t,hi,hf}  \le Cap_{s} \cdot x_{i,d,s,t,hi,hf} \nonumber \qquad 
\\
\forall d \in D \quad
\forall i \in I_{d} \quad 
\forall s \in S \quad
\forall t \in T \quad 
\forall hi \in H \quad 
\forall hf \in H
\end{eqnarray}

\subsubsection{Assigns the classroom of a section to variable 'o'}
\begin{eqnarray}
M \cdot o_{i,d,s}  \geq \sum\limits_{t \in T}\sum\limits_{hi \in H}\sum\limits_{hf \in H} x_{i,d,s,t,hi,hf}  \nonumber \qquad 
\\
\forall d \in D \quad
\forall i \in I_{d} \quad
\forall u \in U \quad
\forall s \in S_{u} \quad
\end{eqnarray}

\subsubsection{Ensures single classroom per course section}
\begin{eqnarray}
\sum\limits_{u \in U} \sum\limits_{s \in S_{u}} o_{i,d,s}  \leq  z_{i,d,cp}  \nonumber \qquad 
\forall d \in D \quad
\forall i \in I_{d} \quad
\end{eqnarray}

\subsubsection{No overlapping in classroom's timetable}
\begin{eqnarray}
\sum\limits_{i \in I_{d}} \sum\limits_{d \in D} \sum\limits_{hi \in H_{d}} \sum_{\substack {hf \in H_{d} \\ hi\le hf}} x_{i,d,s,t,hi,hf}  \leq  1  \nonumber \qquad 
\\
\forall cp \in CP \quad
\forall u \in U_{cp} \quad
\forall s \in S_{u} \quad
\forall t \in T \quad
\forall h \in H \quad (hi,hf)\text{ overlaps }h
\end{eqnarray}

\subsubsection{Tries to satisfy each student requirement}
\begin{eqnarray}
\sum\limits_{i \in I_{d}} s_{i,d,a} + fd_{d,a} = 1  \nonumber \qquad 
\forall a \in Al \quad
\forall d \in D_{a}
\end{eqnarray}

\subsubsection{Assigns student's course section to lessons, while ensures course section's total number of credits}
\begin{eqnarray}
nCreds_{d} \cdot s_{i,d,a} = \sum\limits_{s \in S}\sum\limits_{t \in T}\sum\limits_{hi \in H}\sum\limits_{hf \in H} NHC_{d,hi,hf} \cdot v_{a,i,d,s,t,hi,hf} \nonumber \qquad 
\forall a \in A \quad
\forall i \in I_{d} \quad
\forall d \in D_{a}
\end{eqnarray}

\subsubsection{Links variables x and variable z, while ensures course section's total number of credits}
\begin{eqnarray}
\sum\limits_{s \in S}\sum\limits_{t \in T}\sum\limits_{hi \in H}\sum\limits_{hf \in H} NHC_{d,hi,hf} \cdot x_{i,d,s,t,hi,hf} = nCreds_{d} \cdot z_{i,d,cp} \nonumber \qquad
\forall i \in I_{d} \quad
\forall d \in D \quad
\forall cp \in CP
\end{eqnarray}

\subsubsection{Avoids classes overlapping at student timetable}
\begin{eqnarray}
\sum\limits_{u \in U_{cp}} \sum\limits_{s \in S_{u}} \sum\limits_{i \in I_{d}} \sum\limits_{d \in D_{a}} \sum\limits_{hi \in H_{d}} \sum_{\substack {hf \in H_{d} \\ hi\le hf \\ (hi,hf)\mbox{ overlaps }h}} v_{a,i,d,s,hi,hf,t}  \leq  1  \nonumber \qquad 
\\
\forall cp \in CP \quad
\forall a \in A \quad
\forall t \in T \quad
\forall h \in H_{d}
\end{eqnarray}

\subsubsection{Ensures student assignment to practical and theoretical credits}
\mbox{If practical and theoretical course sections have MxN relationship:}
\begin{eqnarray}
\sum\limits_{i \in I_{dp}} s_{i,dp,a} = \sum\limits_{ i \in I_{dt} } s_{i,dt,a} \nonumber \qquad 
\forall a \in Al \quad
\forall (dp,dt) \in D_{a} \quad
\end{eqnarray}
\mbox{If practical and theoretical course sections have 1x1 relationship:}
\begin{eqnarray}
s_{i,dp,a} = s_{i,dt,a} \nonumber \qquad 
\forall a \in Al \quad
\forall i \in I_{d} \quad
\forall (dp,dt) \in D_{a} \quad
\end{eqnarray}

\subsubsection{Single lesson for each course section per day}
\begin{eqnarray}
\sum\limits_{u \in U_{cp}} \sum\limits_{s \in S_{u}} \sum\limits_{hi \in H_{d}} \sum_{\substack {hf \in H_{d} \\ hi\le hf}} x_{i,d,s,hi,hf,t}  \leq  1  \nonumber \qquad 
\forall d \in D \quad
\forall i \in I_{d} \quad
\forall t \in T \quad
\forall cp \in CP 
\end{eqnarray}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 					CREDITS SPLIT RULES

\subsubsection{Credits split rules}

\paragraph{Credits split rule for each course section}
\begin{eqnarray}
\sum\limits_{u \in U_{cp}} \sum\limits_{s \in S_{u}} \sum\limits_{hi \in H_{d}} \sum_{\substack {hf \in H_{d} \\ hi\le hf}}
 NCH_{d,hi,hf} \cdot x_{i,d,u,s,hi,hf,t} = \sum\limits_{k \in K_{d}}N_{d,k,t} \cdot m_{d,i,k} + fkp_{d,i,t} - fkm_{d,i,t} \nonumber \qquad 
\\
\forall cp \in CP \quad
\forall d \in D \quad
\forall i \in I_{d} \quad
\forall t \in T
\end{eqnarray}

\paragraph{Single credits split rule for each course section}
\begin{eqnarray}
\sum\limits_{k \in K_{d}} m_{d,i,k} \leq 1 \nonumber \qquad 
\forall d \in D \quad
\forall i \in I_{d}
\end{eqnarray}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 					ALOCAÇÃO DE PROFESSORES
\subsubsection{Professors assignment}

\paragraph{No overlapping in professor's timetable}
\begin{eqnarray}
\sum_{ \substack {dti,dtf \in Dt \\ dt \in [dti,dtf)} } \sum\limits_{i} \sum\limits_{d} k_{p,i,d,t,dti,dtf} \le 1 \nonumber \qquad
\forall p \in P \quad
\forall t \in T \quad
\forall dt \in Dt
\end{eqnarray}

\paragraph{Assigns professor to class (sets variable k) - 1}
\begin{eqnarray}
\sum_{ \substack {hi,hf \in H \\ dti \in [hi,hf)} } \sum\limits_{s} x_{i,d,s,t,hi,hf} \le k_{p,i,d,cp,t,dti} + ( 1 - y_{p,i,d,cp} ) \nonumber \qquad
\\
\forall i \in I \quad
\forall d \in D \quad
\forall cp \in CP \quad
\forall p \in P \quad
\forall t \in T \quad
\forall dti \in Dt
\end{eqnarray}
	
\paragraph{Assigns professor to class (sets variable k) - 2}
\begin{eqnarray}
\sum_{ \substack {hi,hf \in H \\ dti \in [hi,hf)} } \sum\limits_{s} x_{i,d,s,t,hi,hf} \le \sum\limits_{p} k_{p,i,d,cp,t,dti} \nonumber \qquad
\\
\forall i \in I \quad
\forall d \in D \quad
\forall cp \in CP \quad
\forall t \in T \quad
\forall dti \in Dt
\end{eqnarray}
	
\paragraph{Assigns professor to class (sets variable y)}
\begin{eqnarray}
\sum\limits_{t} \sum\limits_{h} k_{p,i,d,cp,t,h} = nCreds_{d} \cdot y_{p,i,d,cp} \nonumber \qquad
\forall i \in I \quad
\forall d \in D \quad
\forall cp \in CP \quad
\forall p \in P
\end{eqnarray}	
	
\paragraph{Assigns a single professor to each course section}
\begin{eqnarray}
\sum\limits_{p} y_{p,i,d,cp} = z_{i,d,cp} \nonumber \qquad
\forall i \in I \quad
\forall d \in D \quad
\forall cp \in CP
\end{eqnarray}	


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%						GAP NOS HORÁRIOS DOS PROFESSORES
\subsubsection{Professor's gaps avoidance}

\paragraph{Sets variable $hip_{p,t,f}$}
\begin{eqnarray}
hip_{p,t,f} \geq m(dt) \cdot ( 1 - \sum\limits_{k \in K_{dti<dt}} k_{p,t,dti} ) \nonumber \qquad
\forall p \in P \quad
\forall t \in T \quad
\forall f \in F \quad
\forall dt \in Dt_{f}
\end{eqnarray}
\begin{eqnarray}
hip_{p,t,f} \leq m(dt) + M \cdot ( 1 - \sum\limits_{k} k_{p,t,dti} ) \nonumber \qquad
\forall p \in P \quad
\forall t \in T \quad
\forall f \in F \quad
\forall dti \in Dt_{f}
\end{eqnarray}

\paragraph{Sets variable $hfp_{p,t,f}$}
\begin{eqnarray}
hfp_{p,t,f} \geq \sum\limits_{k} m(dt) \cdot k_{p,t,dtf} \nonumber \qquad
\forall p \in P \quad
\forall t \in T \quad
\forall f \in F \quad
\forall dtf \in Dt_{f}
\end{eqnarray}
\begin{eqnarray}
hfp_{p,t,f} \leq m(dt) + M \cdot ( \sum\limits_{k \in K_{dtf>dt}} k_{p,t,dtf} ) \nonumber \qquad
\forall p \in P \quad
\forall t \in T \quad
\forall f \in F \quad
\forall dt \in Dt_{f}
\end{eqnarray}

\paragraph{Prohibits gap in each session of day in professor timetable}
\begin{eqnarray}
\sum\limits_{k \in K_{h \in H_{f}}} tempo_{k} \cdot k_{p,t,h} + delta_{f,t} + fpgap_{p,t,f} \geq hfp_{p,t,f} - hip_{p,t,f} \nonumber \qquad
\\
\forall p \in P \quad
\forall t \in T \quad
\forall f \in F \quad
\end{eqnarray}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 				GAP NOS HORÁRIOS DOS ALUNOS
\subsubsection{Student's gaps prohibition}

\paragraph{Sets variable $hia_{a,t}$}
\begin{eqnarray}
hia_{a,t} \geq m(dt) \cdot ( 1 - \sum\limits_{v \in V_{dti<dt}} v_{a,t,dti} ) \nonumber \qquad
\forall p \in P \quad
\forall t \in T \quad
\forall dt \in Dt_{f}
\end{eqnarray}
\begin{eqnarray}
hia_{a,t} \leq m(dt) + M \cdot ( 1 - \sum\limits_{v} v_{a,t,dti} ) \nonumber \qquad
\forall p \in P \quad
\forall t \in T \quad
\forall dti \in Dt
\end{eqnarray}

\paragraph{Sets variable $hfa_{a,t}$}
\begin{eqnarray}
hfp_{p,t,f} \geq \sum\limits_{v} m(dt) \cdot v_{a,t,dtf} \nonumber \qquad
\forall p \in P \quad
\forall t \in T \quad
\forall dtf \in Dt
\end{eqnarray}
\begin{eqnarray}
hfa_{a,t} \leq m(dt) + M \cdot ( \sum\limits_{v \in V_{dtf>dt}} v_{a,t,dtf} ) \nonumber \qquad
\forall p \in P \quad
\forall t \in T \quad
\forall dt \in Dt
\end{eqnarray}

\paragraph{Prohibits gap in student timetable}
\begin{eqnarray}
\sum\limits_{i} \sum\limits_{d} \sum\limits_{s} \sum\limits_{hi} \sum\limits_{hf} tempo_{hi,hf} \cdot v_{a,i,d,s,t,hi,hf} + delta_{a,t} \geq hfa_{a,t} - hia_{a,t} \nonumber \qquad
\\
\forall a \in A \quad
\forall t \in T \quad
\end{eqnarray}





\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%    OPERATIONAL MODULE     %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Operational module}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Notation}

\subsubsection{Additional Sets}
\begin{itemize}
\item $I_{d}$ - set of sections for a course $d$.  Elements of $I_{d}$ are called $i$.
\item $I_{d,c}$ - set of sections for a course $d$ that contain student of major $c$. Elements of $I_{d,c}$ are called $i$.
\item $A$ - set of lessons. A lesson is a number of credits of a section $i$ of course $d$, held at some classroom $s$ at a day $t$. Elements of $A$ are called $a$.
\item $P$ - Set of all professors. Elements of $P$ are called $p$.
\item $P_{m}$ - Set of master professors. Professors with higher degree are not included here. Elements of $P_{m}$ are called $p$.
\item $P_{d}$ - Set of doctor professors. Elements of $P$ are called $p$.
\item $PV$ - Set of virtual professors. Elements of $P$ are called $p$.
\item $PV_{d}$ - Set of virtual doctor professors. Elements of $P$ are called $p$.
\item $PV_{m}$ - Set of virtual master professors. Elements of $P$ are called $p$.
\item $PV_{g}$ - Set of virtual not master neither doctor professors. Elements of $P$ are called $p$.
\item $Al$ - Set of virtual student. Elements of $Al$ are called $al$.
\item $R$ - Set of employment agreement types. They are usually full-time, partial-time and hour-paid-time. Elements of $R$ are called $r$.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsubsection{Model data}
\begin{itemize}
\item $m(h)$ - clock time of time slot h in minutes.
\item $minCont_{c,r}$ - minimum percentage of professors with employment agreement type $r$ at major $c$.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Variables}
\begin{itemize}
\item $x_{p,a,h}$ - binary variable that indicates if professor $p$ is assigned to lesson $a$ at time slot $h$.
\item $y_{p,d,i,cp}$ - binary variable that indicates if professor $p$ is assigned to section $i$ of course $d$.
\item $w_{p,c}$ - binary variable that indicates if professor $p$ teaches classes for major $c$.
\item $fmm_{c}$ - integer slack variable for the minimum percentage of master professors in major $c$.
\item $fmd_{c}$ - integer slack variable for the minimum percentage of doctor professors in major $c$.
\item $fchmp_{p}$ - integer slack variable for the minimum workload for professor $p$ when compared with the previous semester.
\item $fchip_{p}$ - integer slack variable for the minimum workload for professor $p$.
\item $fchsp_{p}$ - integer slack variable for the maximum workload for professor $p$
\item $pt_{p,t}$ - binary variable that indicates if professor $p$ teaches at day $t$.
\item $np_{c}$ - integer variable that indicates the number of non-virtual professors assigned to major $c$.
\item $npv_{c,r}$ - integer variable that indicates the number of virtual professors with employment agreement $r$ assigned to major $c$.
\item $npvm_{c,r}$ - integer variable that indicates the number of virtual master professors with employment agreement $r$ assigned to major $c$.
\item $npvd_{c,r}$ - integer variable that indicates the number of virtual doctor professors with employment agreement $r$ assigned to major $c$.
\item $npvg_{c,r}$ - integer variable that indicates the number of virtual not master neither doctor professors with employment agreement $r$ assigned to major $c$.
\item $hip_{p,t,f}$ - integer variable, indicates the time in minutes of the first time slot assigned to professor $p$ at phase $f$ of day $t$. For example, if the first class assigned to $p$ on $t=Monday$ and $f=morning$ starts at 9:30 am, then $hip_{p,t,f}=9\cdot 60 + 30 = 570$.
\item $hfp_{p,t,f}$ - integer variable, indicates the time in minutes of the end of the last time slot assigned to professor $p$ at phase $f$ of day $t$.
\item $fpgap_{p,t,f}$ - integer slack variable, indicates the gap in phase $f$ of day $t$ of professor $p$'s schedule.
\end{itemize}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Objective function}

$$
\begin{array}{rl}
   \mbox{MIN} &
       \sum\limits_{pv \in PV} \sum\limits_{a \in A} \sum\limits_{h \in H} x_{pv,a,h}
      \\
      &
       + \sum\limits_{p \in P} \sum\limits_{i \in I_{d}} \sum\limits_{d \in D}
       ( ( 10 - grade ) \cdot wGrade + ( 10 - pref ) \cdot wPref + costCred \cdot nCredsCourse ) \cdot y_{p,d,i}
      \\
      &
       + \sum\limits_{p \in P} fchmp_{p}
      \\                 
      &
       + \sum\limits_{p \in P} fchip_{p}
      \\
      &
       + \sum\limits_{p \in P} fchsp_{p}
      \\
      &
       + \sum\limits_{c \in C} \sum\limits_{r \in R} npv_{c,r}
      \\   
      &
       + \sum\limits_{c \in C} \sum\limits_{r \in R} npvm_{c,r}
      \\   
      &
       + \sum\limits_{c \in C} \sum\limits_{r \in R} npvd_{c,r}
      \\               
      &
       + \sum\limits_{c \in C} \sum\limits_{r \in R} npvg_{c,r}
      \\                     
      &
       (+-) \sum\limits_{p \in P} \sum\limits_{t \in T} pt_{p,t}
      \\   
      &
      + \sum\limits_{p \in P} \sum\limits_{t \in T} \sum\limits_{f \in F} fpgap_{p,t,f}
\end{array}
$$



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{MIP - Assignment Formulation - Both steps}

\subsubsection{Complete assignment of classes}
\begin{eqnarray}
\sum\limits_{p \in P \cup PV}\sum\limits_{h \in H} x_{p,a,s,d,i,t,h} =  1 \nonumber \qquad 
\forall a \in A 
\end{eqnarray}

\subsubsection{No overlapping in classroom's timetable}
\begin{eqnarray}
\sum\limits_{p \in P \cup PV} \sum\limits_{a \in A} \sum\limits_{\substack {h' \in H \\ h'\mbox{ overlaps }h}} x_{p,a,h'} \leq  1 \nonumber \qquad 
\forall s \in S \quad
\forall t \in T \quad
\forall h \in H
\end{eqnarray}

\subsubsection{No overlapping in professor's timetable}
\begin{eqnarray}
\sum\limits_{a \in A} \sum\limits_{\substack {h' \in H \\ h'\mbox{ overlaps }h}} x_{p,a,h'} \leq  1 \nonumber \qquad 
\forall p \in P \quad
\forall t \in T \quad
\forall h \in H
\end{eqnarray}

\subsubsection{No overlapping in student's timetable}
\begin{eqnarray}
\sum\limits_{p \in P \cup PV} \sum\limits_{a \in A} \sum\limits_{\substack {h' \in H \\ h'\mbox{ overlaps }h}}
  x_{p,a,h'} \leq  1 \nonumber \qquad 
\forall al \in Al \quad
\forall t \in T \quad
\forall h \in H
\end{eqnarray}

\subsubsection{Links professor to lessons and course sections}
\begin{eqnarray}
\sum\limits_{a \in A} \sum\limits_{s \in S} \sum\limits_{t \in T} \sum\limits_{h \in H} x_{p,a,s,d,i,t,h} \leq  M \cdot y_{p,d,i,cp} \nonumber \qquad 
\forall p \in P \cup PV \quad
\forall cp \in Cp \quad
\forall d \in D \quad
\forall i \in I_{d}
\end{eqnarray}

\subsubsection{Single professor per course section}
\begin{eqnarray}
\sum\limits_{p \in P \cup PV} y_{p,d,i,cp} \leq 1 \nonumber \qquad
\forall cp \in Cp \quad
\forall d \in D \quad
\forall i \in I_{d}
\end{eqnarray}

\subsubsection{Counts the number of real professors per course}
\begin{eqnarray}
\sum\limits_{p \in P} w_{p,c} = np_{c} \nonumber \qquad 
\forall c \in C
\end{eqnarray}

\subsubsection{Ensures the maximum percentage of non-master professors}
\begin{eqnarray}
\sum\limits_{r \in R} npvg_{c,r} + \sum\limits_{p \in P_{g}} w_{p,c} \le (1-mm) \cdot (np_{c} + \sum\limits_{r \in R} npv_{c,r}) \nonumber \qquad 
\forall c \in C \quad
\end{eqnarray}

\subsubsection{Ensures the maximum percentage of non-doctor professors}
\begin{eqnarray}
\sum\limits_{r \in R}( npvg_{c,r} + npvm_{c,r} ) \sum\limits_{p \in P_{g} \cup P_{m} } w_{p,c} \le (1-md) \cdot (np_{c} + \sum\limits_{r \in R} npv_{c,r}) \nonumber \qquad 
\forall c \in C \quad
\end{eqnarray}

\subsubsection{Ensures the minimum percentage of master professors per major}
\begin{eqnarray}
\sum\limits_{r \in R} (npvm_{c,r} + npvd_{c,r}) + \sum\limits_{p \in P_{m \cup d}} w_{p,c} \ge mm_{c} \cdot (np_{c} + \sum\limits_{r \in R} npv_{c,r}) \nonumber \qquad 
\forall c \in C \quad
\end{eqnarray}

\subsubsection{Ensures the minimum percentage of doctor professors per major}
\begin{eqnarray}
\sum\limits_{r \in R} npvd_{c,r} + \sum\limits_{p \in P_{d}} w_{p,c} \ge md_{c} \cdot (np_{c} + \sum\limits_{r \in R} npv_{c,r}) \nonumber \qquad 
\forall c \in C \quad
\end{eqnarray}

\subsubsection{Ensures the maximum percentage of professors without a employment contract type}
\begin{eqnarray}
\sum\limits_{r' \in R_{\le r}}( npv_{c,r'} ) \sum\limits_{p \in P_{\le r} } w_{p,c} \le (1-minCont_{c,r}) \cdot (np_{c} + \sum\limits_{r' \in R} npv_{c,r'}) \nonumber \qquad 
\forall c \in C \quad
\forall r \in R
\end{eqnarray}

\subsubsection{Ensures the minimum percentage of professors with a employment contract type per major}
\begin{eqnarray}
\sum\limits_{r' \in R_{\ge r}} npv_{c,r'} + \sum\limits_{p \in P_{\ge r}} w_{p,c} \ge minCont_{c,r} \cdot (np_{c} + \sum\limits_{r' \in R} npv_{c,r'}) \nonumber \qquad 
\forall c \in C \quad
\forall r \in R
\end{eqnarray}

\subsubsection{Consistence for the number of virtual professors}
\begin{eqnarray}
npvg_{c,r} + npvm_{c,r} + npvd_{c,r} = npv_{c,r} \nonumber \qquad 
\forall c \in C \quad
\end{eqnarray}

\subsubsection{Calculates whether a professor is assigned to some lesson of a major (Sets $w_{p,c}$)}
\begin{eqnarray}
\sum\limits_{d \in D} \sum\limits_{i \in I_{d,c}} y_{p,d,i} \le nSections_{d} \cdot w_{p,c} \nonumber \qquad 
\forall p \in P \quad
\forall c \in C
\end{eqnarray}
\begin{eqnarray}
\sum\limits_{d \in D} \sum\limits_{i \in I_{d,c}} y_{p,d,i} \ge w_{p,c} \nonumber \qquad 
\forall p \in P \quad
\forall c \in C
\end{eqnarray}

\subsubsection{Ensures the minimum workload of a professor when compared with his previous semester's workload}
\begin{eqnarray}
\sum\limits_{i \in I_{d}}\sum\limits_{d \in D} nCreds_{d} \cdot y_{p,d,i} + fchm_{p} \ge chMinPrevious_{p} \nonumber \qquad 
\forall p \in P \quad
\end{eqnarray}

\subsubsection{Ensures the minimum workload of a professor}
\begin{eqnarray}
\sum\limits_{a \in A}\sum\limits_{h \in H} nCreds_{a} \cdot x_{p,a,h} + fchi_{p} \ge chMin_{p} \nonumber \qquad 
\forall p \in P \quad
\end{eqnarray}

\subsubsection{Ensures the maximum workload of a professor}
\begin{eqnarray}
\sum\limits_{i \in I}\sum\limits_{d \in D}\sum\limits_{cp \in CP} nCreds_{d} \cdot y_{p,i,d,cp} \le chMax_{p} \nonumber \qquad 
\forall p \in P \quad
\end{eqnarray}

\subsubsection{Ensures professor's minimum rest time}
\begin{eqnarray}
\sum\limits_{a \in A_{t}} x_{p,a,hf} + \sum\limits_{a \in A_{t+1}} x_{p,a,hi} \le 1 \nonumber \qquad 
\forall p \in P \quad
\forall t \in T \\
\forall \mbox{ (hf,t) and (hi,t+1) s.t. } \Delta_{(hf,t), (hi,t+1)} < MinRestTime
\end{eqnarray}

	
	
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%						GAP NOS HORÁRIOS DOS PROFESSORES
\subsubsection{Professor's gaps avoidance}
	
\paragraph{Sets variable $hip_{p,t,f}$}
\begin{eqnarray}
hip_{p,t,f} \geq m(dt) \cdot ( 1 - \sum\limits_{x \in X_{dti<dt}} x_{p,t,dti} ) \nonumber \qquad
\forall p \in P \quad
\forall t \in T \quad
\forall f \in F \quad
\forall dt \in Dt_{f}
\end{eqnarray}
\begin{eqnarray}
hip_{p,t,f} \leq m(dt) + M \cdot ( 1 - \sum\limits_{x} x_{p,t,dti} ) \nonumber \qquad
\forall p \in P \quad
\forall t \in T \quad
\forall f \in F \quad
\forall dti \in Dt_{f}
\end{eqnarray}

\paragraph{Sets variable $hfp_{p,t,f}$}
\begin{eqnarray}
hfp_{p,t,f} \geq \sum\limits_{x} m(dt) \cdot x_{p,t,dtf} \nonumber \qquad
\forall p \in P \quad
\forall t \in T \quad
\forall f \in F \quad
\forall dtf \in Dt_{f}
\end{eqnarray}
\begin{eqnarray}
hfp_{p,t,f} \leq m(dt) + M \cdot ( \sum\limits_{x \in X_{dtf>dt}} x_{p,t,dtf} ) \nonumber \qquad
\forall p \in P \quad
\forall t \in T \quad
\forall f \in F \quad
\forall dt \in Dt_{f}
\end{eqnarray}

\paragraph{Avoid gap in each session of day in professor timetable}
\begin{eqnarray}
\sum\limits_{x \in X_{h \in H_{f}}} tempo_{x} \cdot x_{p,t,h} + delta_{f,t} + fagap_{a,t} \geq hfp_{p,t,f} - hip_{p,t,f} \nonumber \qquad
\forall p \in P \quad
\forall t \in T \quad
\forall f \in F
\end{eqnarray}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{MIP - Assignment Formulation - Only step 1}

\subsubsection{(R1) Estimates the number of virtual professors per major}
\begin{eqnarray}
\sum\limits_{pv \in PV_{r}}\sum\limits_{a \in A_{c}}\sum\limits_{\substack {h' \in H \\ h'\mbox{ overlaps }h}} x_{pv,a,h'} \le npv_{c,r} \nonumber \qquad 
\forall h \in H \quad
\forall c \in C \quad
\forall r \in R \quad
\forall t \in T
\end{eqnarray}

\subsubsection{(R1) Estimates the number of virtual master professors per major}
\begin{eqnarray}
\sum\limits_{pv \in PV_{m,r}}\sum\limits_{a \in A_{c}}\sum\limits_{\substack {h' \in H \\ h'\mbox{ overlaps }h}} x_{pv,a,h'} \le npvm_{c,r} \nonumber \qquad 
\forall h \in H \quad
\forall c \in C \quad
\forall r \in R \quad
\forall t \in T
\end{eqnarray}

\subsubsection{(R1) Estimates the number of virtual doctor professors per major}
\begin{eqnarray}
\sum\limits_{pv \in PV_{d,r}}\sum\limits_{a \in A_{c}}\sum\limits_{\substack {h' \in H \\ h'\mbox{ overlaps }h}} x_{pv,a,h'} \le npvd_{c,r} \nonumber \qquad 
\forall h \in H \quad
\forall c \in C \quad
\forall r \in R \quad
\forall t \in T
\end{eqnarray}

\subsubsection{(R1) Estimates the number of virtual general professors per major}
\begin{eqnarray}
\sum\limits_{pv \in PV_{g,r}}\sum\limits_{a \in A_{c}}\sum\limits_{\substack {h' \in H \\ h'\mbox{ overlaps }h}} x_{pv,a,h'} \le npvg_{c,r} \nonumber \qquad 
\forall h \in H \quad
\forall c \in C \quad
\forall r \in R \quad
\forall t \in T
\end{eqnarray}

\subsubsection{(R1) Ensures minimum number of professors per major. It is needed when there is a non-trivial maximum number of courses assigned to a professor per course.}
\begin{eqnarray}
MaxDiscProfCurso_{c} \cdot (np_{c} + \sum\limits_{r \in R} npv_{c,r}) \ge qtdDiscComAulas_{c} \nonumber \qquad 
\forall c \in C \quad
\end{eqnarray}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{MIP - Assignment Formulation - Only step 2}

\subsubsection{(R2) Calculates the number of virtual professors per course}
\begin{eqnarray}
\sum\limits_{p \in PV_{r}} w_{p,c} = npv_{c,r} \nonumber \qquad 
\forall c \in C \quad
\forall r \in R
\end{eqnarray}

\subsubsection{(R2) Calculates the number of virtual master professors per course}
\begin{eqnarray}
\sum\limits_{p \in PV_{m,r}} w_{p,c} = npvm_{c,r} \nonumber \qquad 
\forall c \in C \quad
\forall r \in R
\end{eqnarray}

\subsubsection{(R2) Calculates the number of virtual doctor professors per course}
\begin{eqnarray}
\sum\limits_{p \in PV_{d,r}} w_{p,c} = npvd_{c,r} \nonumber \qquad 
\forall c \in C \quad
\forall r \in R
\end{eqnarray}

\subsubsection{(R2) Calculates the number of virtual general professors per course}
\begin{eqnarray}
\sum\limits_{p \in PV_{g,r}} w_{p,c} = npvg_{c,r} \nonumber \qquad 
\forall c \in C \quad
\forall r \in R
\end{eqnarray}
      
\subsubsection{(R2) Virtual professor usage}
\begin{eqnarray}
M \cdot pv_{p} \ge \sum\limits_{c \in C} w_{p,c} \nonumber \qquad 
\forall p \in PV
\end{eqnarray}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
